`respR` - An R package for the analysis of respirometry data

### 2017-11-17 manuscript/LaTeX/manuscript.Rmd

4. Although designed specifically with aquatic respirometry in mind, the object-oriented approach of the package and the dimensionless nature of its analytical functions mean that parts of the package can easily be used to estimate linear relationships from a range of applications in many research disciplines."

@@

opts_chunk$set(collapse = TRUE, comment = "#>", tidy = TRUE, cache = TRUE, 
  highlight = TRUE)
```

# Introduction
Metabolic rate is a fundamental trait associated with virtually all biological functions and is key in predicting patterns in ecology and conservation biology, from populations [@seibel_rate_2007; @barneche_scaling_2014] to ecosystems [@brown_toward_2004]. It has been widely investigated in studies of the effects of external stressors on organisms, and is increasingly used to study resilience in climate change-related studies where warming is expected to drive increased metabolism in ectotherms [e.g. @portner_climate_2002; @carey_sea_2016; @delorme_effects_2016]. By far the most common metric used to assess physiological performance in animals, metabolic rate has been determined from whole organisms to the level of cells and tissues [@white_determinants_2013]. Metabolic rate is typically quantified using respirometry, which measures the rate of oxygen uptake over time, $RO_2$ [@lighton_measuring_2008].

@@

Respirometry studies are increasingly outputting large, high-resolution datasets  conducted over longer periods [e.g. 20 h ; @norin_intraspecific_2012]. In most cases, processing the data involves an *ad hoc* selection of data points with poorly reported criteria, and subsequent manual processing of the data subset(s) using a spreadsheet program (e.g. Microsoft Excel) or an integrated development environment (IDE, e.g. R or Matlab). These approaches can be tedious and time consuming especially when spreadsheet programs struggle with the vast datasets that are generated, while IDEs require a degree of expertise to use and have substantial learning curves. Dedicated software are also available to perform metabolic rate analyses, but many have costly licensing requirements (e.g. AutoResp by Loligo Systems), complicating or preventing their use on multiple machines, and are proprietary and closed-source, hindering scientific reproducibility and transparency.

A number of open-source R software packages have recently become available which are designed for, or are suitable to analyse respirometry data. The "`respirometry`"" package (https://CRAN.R-project.org/package=respirometry) contains a comprehensive collection of tools to explore and evaluate experimental parameters in aquatic respirometry and is useful for the design and diagnosis of experimental setups. The "`rMR`"" package (https://CRAN.R-project.org/package=rMR) performs interval-based metabolic rate calculations and has a method to automatically detect $P_{crit}$ using the "broken-stick" regression method (Yeager & Ultsch, 1989). The package "`LoLinR`"" [@olito_estimating_2017], while not specifically coded for respirometry, provides a statistically robust method of detecting a "best-fit" regression, and performs well at identifying truly linear subsets of a data series. Importantly, all of these packages assume that data are monotonic with evenly-spaced sampling periods with respect to time, an ideal structure but one where respirometry data often fails to meet. They also do not have the tools to analyse complicated datasets to automatically determine useful metric such as $MO_{2,max}$ and $MO_{2,min}$.

@@

Typically, the analysis of respirometry data requires the selection one or more linear sections of the data for more accurate calculations of $RO_2$ [e.g. @gordon_energetics_1989; @chu_ecophysiological_2017; @carey_sea_2016]. The function `calc_rate()` can manually extract and process data subset from respiromety data. Data subsets can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accommodate most, if not all, manual data selection requirements and allow for consistent reporting of methods and results. Rates are determined using linear regression, however a two-point analysis is also included alongside all regression results. Generally, $RO_2$ is calculated using the equation: $$RO_2 = \Delta O_2V$$ where $\Delta O_2$ is the slope of the regression that describes the rate of change in $O_2$ concentration over time, or in the case of a two-point measurement, the difference in $O_2$ concentration divided by time elapsed, and $V$ is the volume of the container [@lighton_measuring_2008].

@@

Units of measurement are not used when determining $RO_2$ in `respR`. Instead, two functions, `convert_DO()` and `convert_rate()`, are used to transform $RO_2$ to specific units and perform volumetric or mass-specific conversions . This unique approach has the advantage of allowing "dimensionless" rate calculations for data that may not be fully supported by the entire respR workflow. For example, measurements of respiration can be scaled to surface area [e.g. diatoms & coral, @iversen_temperature_2013; @naumann_species-specific_2014], volume [e.g. copepod eggs, @hammervold_new_2015], or density [e.g. sea urchin larvae, @stumpp_co2_2011], all of which are currently not supported in `respR`. Users may also be interested in using respR's analytical functions to investigate other linear relationships or perform breakpoint analyses on non-respirometry-related data.

>>>>

4. Although designed specifically with aquatic respirometry in mind, the object-oriented approach of the package and the dimensionless nature of its analytical functions mean that parts of the package can easily be used to estimate any other linear relationships in other research disciplines."

@@

opts_chunk$set(collapse = TRUE, comment = "#>", tidy = TRUE, highlight = TRUE)
```

# Introduction
Metabolic rate is a fundamental trait associated with virtually all biological functions and is key in predicting patterns in ecology and conservation biology, from populations [@seibel_rate_2007; @barneche_scaling_2014] to ecosystems [@brown_toward_2004]. It has been widely investigated in studies of the effects of external stressors on organisms, and is increasingly used to study resilience in climate change-related studies where warming is expected to drive increased metabolism in ectotherms [e.g. @portner_climate_2002; @mcelroy_respiratory_2012; @carey_sea_2016; @delorme_effects_2016]. By far the most common metric used to assess physiological performance in animals, metabolic rate has been determined from whole organisms to the level of cells and tissues [@white_determinants_2013]. Metabolic rate is typically quantified using respirometry, which measures the rate of oxygen uptake over time, $RO_2$ [@lighton_measuring_2008].

@@

Respirometry studies are increasingly outputting large, high-resolution datasets and conducted over longer periods [e.g. 20 h ; @norin_intraspecific_2012]. In most cases, processing the data involves an *ad hoc* selection of data points with poorly reported criteria, and subsequent manual processing of the data subset(s) using a spreadsheet program (e.g. Microsoft Excel) or an integrated development environment (IDE, e.g. R or Matlab). These approaches can be tedious and time consuming especially when spreadsheet programs struggle with the vast datasets that are generated, while IDEs require a degree of expertise to use and have substantial learning curves. Dedicated software are also available to perform metabolic rate analyses, but many have costly licensing requirements (e.g. AutoResp by Loligo Systems), complicating or preventing their use on multiple machines, and are proprietary and closed-source, hindering scientific reproducibility and transparency.

A number of open-source R software packages have recently become available which are designed for, or are suitable to analyse respirometry data. The `respirometry` package (https://CRAN.R-project.org/package=respirometry) contains a comprehensive collection of tools to explore and evaluate experimental parameters in aquatic respirometry and is useful for the design and diagnosis of experimental setups. The `rMR` package (https://CRAN.R-project.org/package=rMR) performs interval-based metabolic rate calculations and has a method to automatically detect $P_{crit}$ using the "broken-stick" regression method (Yeager & Ultsch, 1989). The package `LoLinR` [@olito_estimating_2017], while not specifically coded for respirometry, provides a statistically robust method of detecting a "best-fit" regression, and performs very well at identifying truly linear subsets of a data series. Importantly, all of these packages assume that data are monotonic with evenly-spaced sampling periods with respect to time, an ideal structure but one where respirometry data often fails to meet in reality. They also do not have the tools to analyse complicated datasets to automatically determine useful metric such as $MO_{2,max}$ and $MO_{2,min}$.

@@

Typically, the analysis respirometry data requires the selection one or more linear sections of the raw data for more accurate calculations of $RO_2$ [e.g. @gordon_energetics_1989; @chu_ecophysiological_2017; @carey_sea_2016]. The function `calc_rate()` can manually extract and process data subset from respiromety data. Data subsets can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accommodate most, if not all, manual data selection requirements and allow for consistent reporting of methods and results. Rates are determined using linear regression, however a two-point analysis is also included alongside all regression results. Generally, $RO_2$ is calculated using the equation: $$RO_2 = \Delta O_2V$$ where $\Delta O_2$ is the slope of the regression that describes the rate of change in $O_2$ concentration over time, or in the case of a two-point measurement, the difference in $O_2$ concentration divided by time elapsed, and $V$ is the volume of the container [@lighton_measuring_2008].

@@

For deliberate reasons, units of measurement are not used when determining $RO_2$ in `respR`. Instead, two functions, `convert_DO()` and `convert_rate()`, are used to transform $RO_2$ to specific units and perform volumetric or mass-specific conversions . This unique approach has the advantage of allowing "dimensionless" rate calculations for data that may not be fully supported by the entire respR workflow. For example, measurements of respiration can be scaled to surface area [e.g. diatoms & coral, @iversen_temperature_2013; @naumann_species-specific_2014], volume [e.g. copepod eggs, @hammervold_new_2015], or density [e.g. sea urchin larvae, @stumpp_co2_2011], all of which are currently not supported in `respR`. Users may also be interested in using respR's analytical functions to investigate other linear relationships or perform breakpoint analyses on non-respirometry-related data.
